---
title       : Predicting Interest Rates For Lending Club Loan Applications
subtitle    : Developing Data Products - Coursera - July 2014
author      : Mark H. Butler
job         : student @ Coursera
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Introduction

- Background:
     - The [Lending Club](https://www.lendingclub.com/home.action) is an on-line lending community 
     - It [determines the loan interest rate based on](https://www.lendingclub.com/public/how-we-set-interest-rates.action)
          - characteristics of the loan e.g. amount, period
          - characteristics of the borrower e.g. employment history, credit history
     - Several of the borrower characteristics are used to calculate [FICO credit score](http://en.wikipedia.org/wiki/Credit_score_in_the_United_States)
- Questions:
     - Can we predict the interest rate using [historic data](https://spark-public.s3.amazonaws.com/dataanalysis/loansData.rda) using this [codebook](https://spark-public.s3.amazonaws.com/dataanalysis/loansCodebook.pdf)? 
     - Which variables should be included in the predictive model?
           - Which ones are already included in FICO score?
     - Can we develop an app that predicts the interest rate for prospective credit applications?

--- .class #id 

## ANOVA interest rate and other variables

```{r echo=FALSE, results='asis'}
library(xtable)
load("LendingClubApp/data/loansData.rda")
loansData$FICO.min <- as.numeric(gsub("-(.*)", "", as.character(loansData$FICO.Range)))
loansData$FICO.max <- as.numeric(gsub("(.*)-", "", as.character(loansData$FICO.Range)))
loansData$FICO.mid <- (loansData$FICO.max - loansData$FICO.min) / 2 + loansData$FICO.min
loansData$FICO.min <- NULL
loansData$FICO.max <- NULL
loansData$interest  <- as.numeric(gsub("%","",as.character(loansData$Interest.Rate)))
loansData$debtToIncome  <- as.numeric(gsub("%","",as.character(loansData$Debt.To.Income.Ratio)))
loansData = loansData[complete.cases(loansData),]
loansData$FICO.cut = cut2(loansData$FICO.mid, g=5)
loansData <- droplevels(loansData)
aov.i.out <- aov(loansData$interest ~ 
  loansData$FICO.cut + 
  scale(loansData$Amount.Requested) + 
  loansData$Loan.Length + 
  scale(loansData$Monthly.Income) + 
  loansData$Inquiries.in.the.Last.6.Months + 
  loansData$Loan.Purpose +
  scale(loansData$Open.CREDIT.Lines) + 
  loansData$Home.Ownership  +
  loansData$State +
  scale(loansData$debtToIncome) + 
  scale(loansData$Revolving.CREDIT.Balance) + 
  loansData$Employment.Length + 
  loansData$Inquiries.in.the.Last.6.Months)
print(xtable(aov.i.out), type="html")
```

--- .class #id 

## ANOVA FICO score and other variables

```{r echo=FALSE, results='asis'}
aov.f.out <- aov(loansData$FICO.mid ~ 
  scale(loansData$Monthly.Income) + 
  scale(loansData$debtToIncome) + 
  loansData$Home.Ownership + 
  loansData$Inquiries.in.the.Last.6.Months +
  loansData$Loan.Purpose + 
  scale(loansData$Open.CREDIT.Lines) + 
  loansData$Inquiries.in.the.Last.6.Months + 
  loansData$State + 
  scale(loansData$Revolving.CREDIT.Balance) + 
  loansData$Employment.Length)
print(xtable(aov.f.out), type="html")
```

--- &twocol

## Predictive model

*** =left

- Conclusions: 
   - Several variables are related to the interest rate but some are also related to FICO score
   - So let's use a very simple model using FICO Score, Amount Requested and Loan Length
   - Figure 1 shows the actual interest rate versus predicted interest rate
   - The model achieves an R squared value of `r summary(fit)$r.squared`
   - [Present the predictive model as a Shiny App](http://butlermh.shinyapps.io/LendingClub) 

*** =right

```{r, echo=FALSE}
fit <- lm(interest ~ FICO.cut + scale(log10(Amount.Requested)) + Loan.Length, data=loansData)
plot(loansData$interest, fit$fitted, xlab="Figure 1: Actual interest rate (percent)", ylab="Predicted interest rate (percent)", col=loansData$FICO.cut, main="Actual interest rate versus predicted interest rate for\nmodel using FICO score, Amount Requested and Loan Length")
legend('topleft',col=unique(as.numeric(loansData$FICO.cut))[c(4,2,3,1,5)],legend=unique(loansData$FICO.cut)[c(4,2,3,1,5)],pch=19, title="FICO.cut")
```

